---
title: "CreateEventBoundaries"
author: "Yuchao Wang"
date: "11/7/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Match 2V according to OV, include=FALSE} 
DM2Vmarked <- read.csv(file = "thesis/emoDM2VFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)
for (index in 1:1233) {
  if (is.na(DMOmarked$marker_event[index])) {
    next
  }
  else {
    find.match <- which(DM2Vmarked$word == DMOmarked$word[index], arr.ind = TRUE)
    num.match <- length(find.match)
    if (num.match == 0) {
      print(paste(DMOmarked$word[index], " at row ", index, " has no match in 2V", sep = ""))
    }
    else {
      if (num.match ==1) {
        DM2Vmarked$marker_event[find.match] <- 1 
      }
      else {
        print(paste(DMOmarked$word[index], " at row ", index, " has multiple matches at row ", find.match, " in 2V", sep = ""))
      }
    }
  }
}
#write.csv(DM2Vmarked, file = "thesis/emoDM2VFinal_eventMarked.csv", na = "")

```


## Adding MN and Onset based on event boundaries 

* Event boundaries were created by Franziska Hartung on the original, 1st person perspective version (`O`) of both stories. 
* The same event boundaries were created on the 3rd person perspective versions (`2V`) using a separate script.
* **Do not** run this chunk again because all `_eventMarked.csv`, which initially only contains a column of event markers in addition to the *per word* `Final` files, have been processed to add MN and SE *per event*.    

```{r New columns of event boundaries, include=FALSE} 
emoDHOmarked <- read.csv(file = "emoDHOFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)
emoDH2Vmarked <- read.csv(file = "emoDH2VFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)
emoDMOmarked <- read.csv(file = "emoDMOFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)
emoDM2Vmarked <- read.csv(file = "emoDM2VFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)
litDHOmarked <- read.csv(file = "litDHOFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)
litDH2Vmarked <- read.csv(file = "litDH2VFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)
litDMOmarked <- read.csv(file = "litDMOFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)
litDM2Vmarked <- read.csv(file = "litDM2VFinal_eventMarked.csv", header = T, sep = ",", stringsAsFactors = FALSE)

addEventParam <- function(dataEventMarked) {
#initialize condition for search
  begin_row <- 1
  end_row <- 1
  total_row <- nrow(dataEventMarked)
  #search loop 
  while (end_row <= total_row) {
    if  (is.na(dataEventMarked$marker_event[end_row])) {
      end_row <- end_row+1
    }
    else {
      dataEventMarked$onset_event[end_row] <- dataEventMarked$Onset[begin_row]
      dataEventMarked$duration_event[end_row] <- sum(dataEventMarked$Duration[begin_row:end_row])
      dataEventMarked$MN_event[end_row] <- mean(dataEventMarked$MN[begin_row:end_row], na.rm = TRUE)
      begin_row <- end_row+1
      end_row <- end_row+1
    }
  }
  return(dataEventMarked)
}

emoDHOmarked <- addEventParam(emoDHOmarked)
emoDH2Vmarked <- addEventParam(emoDH2Vmarked)
emoDMOmarked <- addEventParam(emoDMOmarked)
emoDM2Vmarked <- addEventParam(emoDM2Vmarked)
litDHOmarked <- addEventParam(litDHOmarked)
litDH2Vmarked <- addEventParam(litDH2Vmarked)
litDMOmarked <- addEventParam(litDMOmarked)
litDM2Vmarked <- addEventParam(litDM2Vmarked)

write.csv(emoDHOmarked, file = "emoDHOFinal_eventMarked.csv", na = "")
write.csv(emoDH2Vmarked, file = "emoDH2VFinal_eventMarked.csv", na = "")
write.csv(emoDMOmarked, file = "emoDMOFinal_eventMarked.csv", na = "")
write.csv(emoDM2Vmarked, file = "emoDM2VFinal_eventMarked.csv", na = "")
write.csv(litDHOmarked, file = "litDHOFinal_eventMarked.csv", na = "")
write.csv(litDH2Vmarked, file = "litDH2VFinal_eventMarked.csv", na = "")
write.csv(litDMOmarked, file = "litDMOFinal_eventMarked.csv", na = "")
write.csv(litDM2Vmarked, file = "litDM2VFinal_eventMarked.csv", na = "")
```