---
title: "Ratings matching original words"
author: "Yuchao Wang"
date: "8/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
```

```{r, echo = FALSE}
#CLEAN UP DELIMITED CSV FILE CONTAINING ALL RATED WORDS IN SURVEY
template <- read.csv(file = "forWordSeparation_cleaned.csv", header = F, sep = ",", stringsAsFactors = FALSE)
template <- t(template) 
colnames(template) <- template[1, ] 
template <- template[-1, ]
template <- as.data.frame(template) #data type converted to matrix after transpose, due to rectangular shape - force back into data frame for "gather"
template <- gather(template, "qnLabel", "word", 1:17)
template <- template[!(template$word == ""), ] 
template <- template[ , c("word","qnLabel")]
template$qnLabel <- trimws(template$qnLabel)
template$qnLabel <- as.factor(template$qnLabel)
template$qnLabel <- factor(template$qnLabel, levels = c("3.1","5.1","7.1","9.1","11.1","13.1","15.1","17.1","19.1","24.1","26.1","28.1","30.1","32.1","34.1","36.1","38.1")) 
qnLabel <- levels(template$qnLabel)
#------------------------------------------------------- 

#MATCH EMO RESULTS    
emoResult <- read.csv(file="Verhalen_Emotional+Arousal_September+23%2C+2019_08.47.csv", 
                  header = TRUE, sep = ",", stringsAsFactors = FALSE)
emoResult <- emoResult[-c(1,2), ] #get rid of title row
emoResult <- emoResult[-c(1:9), ] #get rid of preview Test results
emoResult <- emoResult[-c(15:17), ] #get rid of incomplete results


totalParticipant <- nrow(emoResult)
totalWord <- nrow(template)
emoResultClean <- template


for (indexResponse in 1:totalParticipant) {
  for (indexWord in 1:totalWord) {
    searchWord <- template[indexWord,1]
    searchQn <- template[indexWord,2]
    resultCol <- indexResponse+2
    emoResultClean[indexWord,resultCol] <- NA #initialize a new column
      for (possibleRating in 1:7) {
        if (!is.na(emoResultClean[indexWord,resultCol])) {
          break
        }
        currentColName <- paste ("Q", searchQn, "_", possibleRating, sep = "", collapse = NULL)
        if (grepl(searchWord, emoResult[indexResponse,currentColName], fixed = TRUE)) {
            emoResultClean[indexWord,resultCol] <- substr(currentColName, nchar(currentColName), nchar(currentColName))        
        }
        else {
          if ((possibleRating == 7) & (is.na(emoResultClean[indexWord,resultCol]))) {
            emoResultClean[indexWord,resultCol] <- NA
          }
        }
      }
  }
}

#------------------------------------------------------- 

#MATCH LIT RESULTS
litResult <- read.csv(file="Verhalen_Literariness_September+23%2C+2019_08.46.csv", 
                  header = TRUE, sep = ",", stringsAsFactors = FALSE)
litResult <- litResult[-c(1,2), ] #get rid of title row
litResult <- litResult[-c(1:10), ] #get rid of preview Test results


totalParticipant <- nrow(litResult)
totalWord <- nrow(template)
litResultClean <- template


for (indexResponse in 1:totalParticipant) {
  for (indexWord in 1:totalWord) {
    searchWord <- template[indexWord,1]
    searchQn <- template[indexWord,2]
    resultCol <- indexResponse+2
    litResultClean[indexWord,resultCol] <- NA #initialize a new column
      for (possibleRating in 1:2) {
        if (!is.na(litResultClean[indexWord,resultCol])) {
          break
        }
        currentColName <- paste ("Q", searchQn, "_", possibleRating, sep = "", collapse = NULL)
        if (grepl(searchWord, litResult[indexResponse,currentColName], fixed = TRUE)) {
            litResultClean[indexWord,resultCol] <- substr(currentColName, nchar(currentColName), nchar(currentColName))        
        }
        else {
          if ((possibleRating == 2) & (is.na(litResultClean[indexWord,resultCol]))) {
            litResultClean[indexWord,resultCol] <- NA
          }
        }
      }
  }
}

#------------------------------------------------------- 
saveRDS(emoResultClean, file = "emoResultClean.rds")
saveRDS(litResultClean, file = "litResultClean.rds")
saveRDS(template, file = "template.rds")

```

```{r} 
#Visualize ratings 
library(ggplot2)
library(psych)
#EMO
temoResultClean <- t(emoResultClean)
temoResultClean <- temoResultClean[-c(1:2),]

emoTrend <- describe(temoResultClean)
emoDMTrend <- emoTrend[c(1:1237),]
emoDHTrend <- emoTrend[c(1238:2354),]
emoDHTrend$vars <- seq(length=nrow(emoDHTrend)) 

dodge <- position_dodge(width = 0.9)
limits <- aes(ymax = emoDMTrend$mean + emoDMTrend$se,
              ymin = emoDMTrend$mean - emoDMTrend$se)

ggplot(data = emoDMTrend, aes(x = vars, y = mean, color = vars)) + 
  geom_point(size=2.5, alpha=0.6, color = "orange") +
  geom_errorbar(limits, position = dodge, width = 0.25) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.x=element_blank()) + 
  ggtitle("Emotional Arousal Rating for De Muur, N=23")
ggsave("~/Desktop/emoDM.png", plot = last_plot(), device = png(),
  scale = 1, width = 300, height = 200, units = c("mm"),
  dpi = 300, limitsize = TRUE)

limits <- aes(ymax = emoDHTrend$mean + emoDHTrend$se,
              ymin = emoDHTrend$mean - emoDHTrend$se)

ggplot(data = emoDHTrend, aes(x = vars, y = mean, color = vars)) + 
  geom_point(size=2.5, alpha=0.6, color = "orange") +
  geom_errorbar(limits, position = dodge, width = 0.25) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.x=element_blank()) + 
  ggtitle("Emotional Arousal Rating for De Mexicaanse Hond, N=23") 
ggsave("~/Desktop/emoDH.png", plot = last_plot(), device = png(),
  scale = 1, width = 300, height = 200, units = c("mm"),
  dpi = 300, limitsize = TRUE)

#LIT
tlitResultClean <- t(litResultClean)
tlitResultClean <- tlitResultClean[-c(1:2),]
litTrend <- describe(tlitResultClean)
litDMTrend <- litTrend[c(1:1237),]
litDHTrend <- litTrend[c(1238:2354),]
litDHTrend$vars <- seq(length=nrow(litDHTrend)) 

dodge <- position_dodge(width = 0.9)
limits <- aes(ymax = litDMTrend$mean + litDMTrend$se,
              ymin = litDMTrend$mean - litDMTrend$se)

ggplot(data = litDMTrend, aes(x = vars, y = mean, color = vars)) + 
  geom_point(size=2.5, alpha=0.6, color = "orange") +
  geom_errorbar(limits, position = dodge, width = 0.25) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.x=element_blank()) + 
  ggtitle("Literariness Rating for De Muur, N=24")
ggsave("~/Desktop/litDM.png", plot = last_plot(), device = png(),
  scale = 1, width = 300, height = 200, units = c("mm"),
  dpi = 300, limitsize = TRUE)

limits <- aes(ymax = litDHTrend$mean + litDHTrend$se,
              ymin = litDHTrend$mean - litDHTrend$se)

ggplot(data = litDHTrend, aes(x = vars, y = mean, color = vars)) + 
  geom_point(size=2.5, alpha=0.6, color = "orange") +
  geom_errorbar(limits, position = dodge, width = 0.25) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.x=element_blank()) + 
  ggtitle("Literariness Rating for De Mexicaanse Hond, N=24")
ggsave("~/Desktop/litDH.png", plot = last_plot(), device = png(),
  scale = 1, width = 300, height = 200, units = c("mm"),
  dpi = 300, limitsize = TRUE)


```
$useless Code---------------------------
dodge <- position_dodge(width = 0.9)
limits <- aes(ymax = litTrend$mean + litTrend$se,
              ymin = litTrend$mean - litTrend$se)

ggplot(data = litTrend, aes(x = vars, y = mean, color = vars)) + 
  geom_point(size=2.5, alpha=0.6, color = "orange") +
  geom_errorbar(limits, position = dodge, width = 0.25) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.x=element_blank()) + 
  ggtitle("Literariness Rating (De Muur + De Hond concatenated), N=24")
----------------------------------------
