---
title: "TMSTT_McRae"
author: "Yuchao Wang"
date: "2/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Matching McRae with MRC
Matching McRae 2005 ratings with additional ratings from MRC 


```{r}
CONCS_brm <- read.delim("~/Documents/GitHub/RStudio/TMSTT/McRae-BRM-InPress/CONCS_brm.txt")

McRae_MRC <- read.delim("~/Documents/GitHub/RStudio/TMSTT/McRae-BRM-InPress/McRae_MRC.txt")
McRae_MRC <- data.frame(lapply(McRae_MRC, function(x) {gsub("-", "NA", x)}))
McRae_MRC[, 2:7] <- sapply(McRae_MRC[, 2:7], as.numeric)


totalMRC <- nrow(McRae_MRC)
totalBRM <- nrow(CONCS_brm)
CombinedList <- cbind(CONCS_brm[FALSE, ], McRae_MRC[FALSE, 2:7]) #create empty dataframe with all relevant variables

for (BRMindex in 1:totalBRM) {
  for (MRCindex in 1:BRMindex) {
    searchString <- paste("^",CONCS_brm$Concept[BRMindex],"$", sep = "") #this makes sure that the strings match exactly, to avoid problems of grep matching e.g. book to bookcase
    if (grepl(searchString, McRae_MRC$Word[MRCindex], ignore.case = TRUE)) {
      CombinedList <- rbind(CombinedList, (cbind(CONCS_brm[BRMindex, ], McRae_MRC[MRCindex, 2:7])))
    }
  }
}

CombinedList_short <- subset(CombinedList, select = c("Concept","ln.KF.", "Familiarity", "Length_Phonemes", "Length_Syllables", "Num_Feats_Tax", "Num_Feats_No_Tax", "Num_Disting_Feats_No_Tax","Density_No_Tax", "CNC", "IMG"))
CombinedList_short <- CombinedList_short[complete.cases(CombinedList_short), ]
write.csv(CombinedList_short, file = "~/Documents/GitHub/RStudio/TMSTT/McRae-BRM-InPress/CombinedList_short.csv", row.names = FALSE)
```

##Check difference in MRC and McRae ratings of familiarity 

Highly correlated (r=0.7), using McRae 2005 since it was more recent than MRC and was done in the US. 

##Check if possible to increase total number of features

Higher familiarity correlates with higher number of features produced 

##Check group differences in number of taxonomic features 

Animals tend to have more taxonomic features than household items and food 

```{r}
cor.test(CombinedList$Familiarity, CombinedList$FAM, method = "pearson")
plot(CombinedList$Familiarity, CombinedList$FAM)

library(ggplot2)
cor.test(CombinedList$Num_Feats_Tax, CombinedList$Familiarity, method = "pearson")
Fam_NumFeats <- ggplot(data = CombinedList, aes(x = Num_Feats_Tax, y = Familiarity, label = Concept)) + 
  geom_text(size = 1.5, alpha = 0.7, angle = 20, position = position_jitter(width = 1, height = 0.5)) +
  xlab("Total Number of Features") +
  ylab("Familiarity") +
  theme(text = element_text(size=15))
ggsave("Fam_NumFeats.png", plot = Fam_NumFeats, device = png(),
    scale = 1, width = 300, height = 180, units = c("mm"),
    dpi = 600, limitsize = TRUE)

library(readxl)
CombinedList_short_grouped <- read_excel("~/Documents/GitHub/RStudio/TMSTT/McRae-BRM-InPress/CombinedList_short_grouped.xlsx")
ggplot(data = CombinedList_short_grouped, aes(x = Group, y = Num_Tax)) +
  geom_boxplot()
```

##Check agreement in 3 lists of stimuli 

- Need 3 lists of 10 words each
- Within each list, 4 household items, 3 animals, 3 food 
- Need to match at least on familiarity, concreteness, and imageability 

```{r}
PotentialList <- read.csv("~/Documents/GitHub/RStudio/TMSTT/McRae-BRM-InPress/Potential_Stim.csv")

summary(manova(cbind(Familiarity, CNC, IMG) ~ ListNum, data = PotentialList)) #one-way manova on familiaritym concreteness, and imageability
summary(manova(cbind(ln.KF., Familiarity, Length_Phonemes, Length_Syllables, Num_Feats_Tax, Num_Feats_No_Tax, Num_Disting_Feats_No_Tax, Density_No_Tax, CNC, IMG) ~ ListNum, data = PotentialList)) #one-way manova on potentially relevant normed features from McRae 2005
```

