---
title: "STAT021 Lecture 21 MLR Review Examples"
fontsize: 12pt
date: "November 27, 2018"
output:
  html_document: default
---

## The Happy Planet Index example

```{r, fig.height=6, fig.width=7.5}
# Input data and exploratory data analysis
load("/home/stat021f18/datasets/HappyPlanet.RData")
head(HappyPlanet)
library(psych)
describe(HappyPlanet)[,2:4]
table(HappyPlanet$HDI)
# If there aren't too many missing observations, we can remove them
# at the beginning of the analysis
HappyPlanet <- HappyPlanet[complete.cases(HappyPlanet), ]
describe(HappyPlanet)[,2:4]
library(GGally)
ggpairs(HappyPlanet[,c("GDPpc","LifeExp","LifeSat","HDI","Happiness")])
library(car)
vif(lm(Happiness ~ GDPpc + LifeExp + LifeSat + HDI, data=HappyPlanet))
# When categorical predictor is in the model, use the "GVIF^(1/(2*Df))"
# values and compare them to sqrt(5) instead of 5.
```

Potential predictors:

* *GDPpc* or *log(GDPpc)*, quadratic terms
* *LifeExp* or *LE2* (categorizing *LifeExp*), quadratic terms for *LifeExp*
* *LifeSat*, quadratic terms
* *HDI*, *HDI* as quantitative
* Two-way interactions

```{r}
# Prepare the potential predictors
logGDPpc <- log(HappyPlanet$GDPpc)
summary(HappyPlanet$LifeExp)
LE2 <- cut(HappyPlanet$LifeExp, breaks=c(30, 70, 90), 
           labels=c("Low","High"))
HDInum <- as.numeric(HappyPlanet$HDI) - 1
HappyPlanet <- data.frame(HappyPlanet, logGDPpc, LE2, HDInum)
```

```{r}
# Zero model and full models
zeroModel <- lm(Happiness ~ 1, data=HappyPlanet)
fullModel1 <- lm(Happiness ~ (GDPpc + LifeExp + LifeSat + HDI)^2
                 + I(GDPpc^2) + I(LifeExp^2) + I(LifeSat^2), data=HappyPlanet)
fullModel2 <- lm(Happiness ~ (logGDPpc + LifeExp + LifeSat + HDI)^2
                 + I(logGDPpc^2) + I(LifeExp^2) + I(LifeSat^2), data=HappyPlanet)
fullModel3 <- lm(Happiness ~ (logGDPpc + LE2 + LifeSat + HDI)^2
                 + I(logGDPpc^2) + I(LifeSat^2), data=HappyPlanet)
fullModel4 <- lm(Happiness ~ (logGDPpc + LifeExp + LifeSat + HDInum)^2
                 + I(logGDPpc^2) + I(LifeExp^2) + I(LifeSat^2), data=HappyPlanet)
```

```{r}
fw1 <- step(zeroModel, scope=list(upper=fullModel1), direction="forward", trace=0)
bw1 <- step(fullModel1, scope=list(lower=zeroModel), direction="backward", trace=0)
st1 <- step(zeroModel, scope=list(upper=fullModel1), direction="both", trace=0)
summary(fw1)
summary(bw1)
summary(st1)
c(summary(fw1)$adj, summary(bw1)$adj, summary(st1)$adj)
anova(fw1, bw1, st1)
anova(fw1, bw1, st1, test="Cp")
AIC(fw1, bw1, st1)
```

Model `bw1` and `st1` are the same, while `fw1` has one additional linear term of *LifeExp*. Since all three models have the quadratic term of *LifeExp*, we choose `fw1` as the best model among the three.

```{r, fig.width=7, fig.height=3}
# Checking assumptions
# Here I write a function to generate the plots and test
library(ggplot2)
library(gridExtra) # You may need to install the gridExtra pacakage first
library(lmtest)
CheckAssumptions <- function(model){
  Assess <- data.frame(Residuals=model$residuals, 
                     FittedValues=model$fitted.values)
  # Residuals vs. Fitted Values
  p1 <- ggplot(data=Assess, aes(x=FittedValues, y=Residuals))+
          geom_point(color="skyblue3", size=2, alpha=0.8)+
          geom_hline(yintercept=0, size=1.2, colour="grey50")+
          ggtitle("Residuals vs. Fitted Values")
  # Normal Q-Q plot
  p2 <- ggplot(data=Assess, aes(sample = scale(Residuals)))+
          stat_qq(size=3, color="skyblue3", alpha=0.7)+
          geom_abline(intercept=0, slope=1, size=1.2, colour="grey50")+
          ggtitle("Normal Q-Q Plot")
  # Histogram of residuals
  p3 <- ggplot(Assess, aes(Residuals))+
          geom_histogram(bins=20, fill="skyblue3", alpha=0.7)+
          ggtitle("Histogram of Residuals")
  grid.arrange(p1, p2, p3, nrow = 1)
  bptest(model)
}
CheckAssumptions(fw1)
```

```{r}
# My function for unusual points
library(MASS)
UnusualPoints <- function(model, data){
  n <- nrow(data) # sample size
  Leverage <- hatvalues(model) # leverage values
  StdRes <- stdres(model) # standardized residuals
  CooksD <- cooks.distance(model) # Cook's D
  unusual <- data.frame(HappyPlanet, Leverage, StdRes, CooksD)
  cd <- function(h, type){sqrt((1-h)/h)*type} # function for cook's D cutoffs
  K <- length(model$coefficients) - 1 # number of predictors
  ggplot(unusual, aes(x=Leverage, y=StdRes, label=Country))+
    geom_point(color="skyblue3", size=2.5, alpha=0.7)+
    geom_vline(xintercept = c(2*(K+1)/n, 3*(K+1)/n), color="red", linetype=2)+
    geom_hline(yintercept = c(-3,-2,2,3), color="orange", linetype=2)+
    stat_function(fun=cd, args=list(type=sqrt(K+1)), color="blue", linetype=2)+
    stat_function(fun=cd, args=list(type=-sqrt(K+1)), color="blue", linetype=2)+
    stat_function(fun=cd, args=list(type=sqrt(0.5*(K+1))), color="blue", linetype=2)+
    stat_function(fun=cd, args=list(type=-sqrt(0.5*(K+1))), color="blue", linetype=2)+
    ggtitle("Diagnostic Plot")
}

UnusualPoints(model=fw1, data=HappyPlanet)
```

```{r}
# Model with logGDPpc instead of GDPpc
fw2 <- step(zeroModel, scope=list(upper=fullModel2), direction="forward", trace=0)
bw2 <- step(fullModel2, scope=list(lower=zeroModel), direction="backward", trace=0)
st2 <- step(zeroModel, scope=list(upper=fullModel2), direction="both", trace=0)
summary(fw2)
summary(bw2)
summary(st2)
c(summary(fw2)$adj, summary(bw2)$adj, summary(st2)$adj)
anova(fw2, bw2, st2)
anova(fw2, bw2, st2, test="Cp")
AIC(fw2, bw2, st2)
```

Model `bw2` and `st2` are the same and seem to be better than `fw2`. But note it does not have the linear term of *LifeExp*. Let's add it in and compare `st2new` to `fw1` to see whether we should use *GDPpc* or *logGDPpc*.

```{r}
st2new <- lm(Happiness ~ LifeSat + HDI * logGDPpc + I(logGDPpc^2)
             + LifeExp + I(LifeExp^2), data=HappyPlanet)
c(summary(fw1)$adj, summary(st2new)$adj)
anova(fw1, st2new)
anova(fw1, st2new, test="Cp")
AIC(fw1, st2new)
```

```{r, fig.width=7, fig.height=3}
CheckAssumptions(st2new)
```

```{r}
UnusualPoints(model=st2new, data=HappyPlanet)
```

Using *logGDPpc* is probably better. But it's totally reasonable to choose `fw1` for it is simpler and not significantly different from `st2new` and has smaller Cp.

```{r}
# Model with LE2 instead of LifeExp
fw3 <- step(zeroModel, scope=list(upper=fullModel3), direction="forward", trace=0)
bw3 <- step(fullModel3, scope=list(lower=zeroModel), direction="backward", trace=0)
st3 <- step(zeroModel, scope=list(upper=fullModel3), direction="both", trace=0)
summary(fw3)
summary(bw3)
summary(st3)
c(summary(fw3)$adj, summary(bw3)$adj, summary(st3)$adj)
anova(fw3, bw3, st3)
anova(fw3, bw3, st3, test="Cp")
AIC(fw3, bw3, st3)
```

These models have much smaller $R^2$ and adjusted $R^2$ than the previous models. We should model *LifeExp* as a quantitative rather than a binary variable.

```{r}
# Model with HDInum instead of HDI
fw4 <- step(zeroModel, scope=list(upper=fullModel4), direction="forward", trace=0)
bw4 <- step(fullModel4, scope=list(lower=zeroModel), direction="backward", trace=0)
st4 <- step(zeroModel, scope=list(upper=fullModel4), direction="both", trace=0)
summary(fw4)
summary(bw4)
summary(st4)
c(summary(fw4)$adj, summary(bw4)$adj, summary(st4)$adj)
anova(fw4, bw4, st4)
anova(fw4, bw4, st4, test="Cp")
AIC(fw4, bw4, st4)
```

Model `bw4` and `st4` are the same and seem to be better than `fw4`. But note it does not have the linear term of *LifeExp*. Let's add it in and compare `st4new` to `fw1` and `st2new`.


```{r}
st4new <- lm(Happiness ~ LifeSat + LifeExp + logGDPpc * HDInum
             + I(LifeExp^2) + I(LifeSat^2), data=HappyPlanet)
c(summary(st1)$adj, summary(st2new)$adj, summary(st4new)$adj)
anova(st1, st2new, st4new)
anova(st1, st2new, st4new, test="Cp")
AIC(st1, st2new, st4new)
```

It turns out `st4new` is the best, where we model *HDI* quantitatively.

Final model analysis

```{r}
describe(HappyPlanet[,c("logGDPpc","LifeExp","LifeSat","HDInum","Happiness")])[,2:4]
ggpairs(HappyPlanet[,c("logGDPpc","LifeExp","LifeSat","HDInum","Happiness")])
vif(lm(Happiness ~ logGDPpc + LifeExp + LifeSat + HDInum, data=HappyPlanet))
```

```{r}
summary(st4new)
```

```{r, fig.width=7, fig.height=3}
CheckAssumptions(st4new)
```

```{r}
UnusualPoints(model=st4new, data=HappyPlanet)
```

```{r}
predict(st4new, list(LifeExp=75, LifeSat=6.9, logGDPpc=log(9223), HDInum=2),
        interval = "confidence")
predict(st4new, list(LifeExp=75, LifeSat=6.9, logGDPpc=log(9223), HDInum=2),
        interval = "prediction")
```